### 2022-12-29 ms.Rnw
minor correction incorporated into accepted stat sinica version

%%% Statistica Sinica formatting
%\newcommand\statSinica[1]{#1}
 \newcommand\statSinica[1]{}
%%% arXiv formatting
 \newcommand\arxiv[1]{#1}
%\newcommand\arxiv[1]{}

@@

The number of travelers from town $\unit$ to $\altUnit$ is denoted by $v_{\unit\altUnit}$.
Here, $v_{\unit\altUnit}$ is constructed using the gravity movement model of \cite{xia04} given by
\[
v_{\unit\altUnit} = \gravity_{\unit} \cdot \frac{\hspace{2mm} \overline{\dist}\hspace{2mm}}{\hspace{1.5mm}\overline{\pop}^2\hspace{1mm}} \cdot \frac{\pop_\unit \cdot \pop_{\altUnit}}{\dist_{\unit\altUnit}},
\]
where $\gravity_u$ is the {\it gravitational constant}, $\dist_{\unit\altUnit}$ is the distance between town $\unit$ and town $\altUnit$, $\pop_\unit$ is the average of $P_{\unit}(t)$ across time, $\overline{\pop}$ is the average of $\pop_{\unit}$ across towns, and $\overline{\dist}$ is the average of $\dist_{\unit\altUnit}$ across all pairs of towns.
The transition processes $N_{EI,\unit}(t)$, $N_{IR,\unit}(t)$ and $N_{QD,\unit}(t)$, for $Q\in\{S,E,I,R\}$,  are modeled as conditional Poisson process with per-capita rates $\mu_{EI,\unit}$, $\mu_{IR,\unit}$ and $\mu_{QD,\unit}$ respectively, and we fix $\mu_{QD,\unit}=(50 \mbox{ year})^{-1}$.
The birth process $N_{BS,\unit}(t)$ is an inhomogeneous Poisson process with rate $\mu_{BS,\unit}(t)$, given by interpolated census data.

@@

The simulated model is drawn from model {\modelA}, which is nested within model {\modelB} but not within {\modelC}.
For models {\modelB} and {\modelC}, with all parameters are unit-specific, we must estimate $20\times 12=240$ parameters.
This greatly exceeds the 7 shared parameters fitted by \citet{ionides21} for a measles metapopulation model using a large computational effort.
An iterated guided intermediate resampling filter (IGIRF) algorithm was used to fit 9 shared parameters and 3 unit-specific initial value parameters in a measles metapopulation model \citep{park20}.
That analysis used a customized treatment for the unit-specific initial value parameters and did not attempt to estimate other unit-specific parameters.
IGIRF is sensitive to the choice of the guide function, and the model-specific implementation used by \citet{park20} performed better than the generic implementation of IGIRF which is currently available in \code{spatPomp}.
To our knowledge, {\ibpf} is therefore advancing the current limits for scalability of simulation-based maximum-likelihood inference for metapopulation dynamics.

@@

<<sim_search_boxplot, echo=F, fig.height=3, fig.width=5.5, out.width="5in", fig.cap = "Fitting simulated measles data, using an initial search and four refinement steps. (A1--A5) model \\modelA, $4\\times 20$ unit-specific parameters and $8$ shared parameters; (B1--B5) model \\modelB, $12\\times 20$ unit-specific parameters and no shared parameters;  (C1--C5) model \\modelC, also all unit-specific, but with immigration rather than coupling, matching He et al (2010). The horizontal dashed line is the log-likelihood at the true parameters, evaluated by BPF.",fig.show = "hold">>=
a <- cbind(
A1 = v1a$loglik,
A2 = v2a$loglik,
A3 = v3a$loglik,
A4 = v4a$loglik,
A5 = v5a$loglik
)

@@

The MLE is not known exactly in this case.
Wilks' theorem gives an asymptotic expectation that the log-likelihood at the MLE should be greater than the likelihood at the truth by approximately $1/2$ the number of parameters, which here is $(4\times 20+8)/2=44$ for {\modelA} and $(12\times 20)/2=120$ for {\modelB}.
The $y$-axis values in Fig.~\ref{fig:sim_search_boxplot} show log-likelihoods exceeding the truth by less than this, indicating some imperfection to the Monte Carlo maximization so far as Wilks' asymptotic result holds in this situation.
Despite this limitation, searches that exceed the likelihood at the truth have found inferentially plausible sets of parameters which can be used to study the likelihood surface around the maximum.
For example, Monte Carlo profile confidence intervals can give proper coverage even in the presence of considerable Monte Carlo error \citep{ionides17,ning21}.

@@

@@

\arxiv{
\setcounter{section}{0}
\setcounter{equation}{0}
\def\theequation{S\arabic{section}.\arabic{equation}}
%\def\thesection{S\arabic{section}}
\renewcommand\thefigure{S\arabic{figure}}
\renewcommand\thetable{S\arabic{table}}
\def\thesection{S\arabic{section}}

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{|l|c|c|c|c|}
\hline
Parameter & \modelA & \modelB & \modelC & Simulation \\
\hline
$\pinit_{S,u}$ & \measlesUnitSpecific & \measlesUnitSpecific & \measlesUnitSpecific &
  $\Sexpr{myround(theta["S_01"],3)}$ \\
$\pinit_{E,u}$ & \measlesUnitSpecific & \measlesUnitSpecific & \measlesUnitSpecific &
  $\Sexpr{myround(theta["E_01"],5)}$ \\
$\pinit_{I,u}$ & \measlesUnitSpecific & \measlesUnitSpecific & \measlesUnitSpecific &
  $\Sexpr{myround(theta["I_01"],5)}$ \\
$\rho_u$ & \measlesUnitSpecific & \measlesUnitSpecific & \measlesUnitSpecific &
  $\Sexpr{myround(theta["rho1"],1)}$ \\
$\measOD_{\unit}$ & \measlesShared & \measlesUnitSpecific & \measlesUnitSpecific &
  $\Sexpr{myround(theta["psi1"],2)}$ \\  
$\mu_{EI,\unit}$ & \measlesShared & \measlesUnitSpecific & \measlesUnitSpecific &
  $\Sexpr{myround(theta["sigma1"]/52,0)} \, \mathrm{week}^{-1}$ \\
$\mu_{IR}$ & \measlesShared & \measlesUnitSpecific & \measlesUnitSpecific &
  $\Sexpr{myround(theta["gamma1"]/52,0)} \, \mathrm{week}^{-1}$ \\
$\meanBeta_\unit$ & \measlesShared & \measlesUnitSpecific & \measlesUnitSpecific &
  $\Sexpr{myround(theta["gamma1"]*(theta["R01"]+theta["mu1"])/52,0)} \, \mathrm{week}^{-1}$ \\
$\sigma_{SE,\unit}$ & \measlesShared & \measlesUnitSpecific & \measlesUnitSpecific &
  $\Sexpr{myround(theta["sigmaSE1"],2)} \, \mathrm{year}^{1/2}$ \\
$\amplitude_{\unit}$ & \measlesShared & \measlesUnitSpecific & \measlesUnitSpecific &
  $\Sexpr{myround(theta["amplitude1"],1)}$ \\
$\alpha_{\unit}$ & \measlesShared & \measlesUnitSpecific & \measlesUnitSpecific &
  $\Sexpr{myround(theta["alpha1"],0)}$ \\
$\gravity_{\unit}$ & \measlesShared & \measlesUnitSpecific & 0 &
  $\Sexpr{myround(theta["g1"],0)}$ \\
$\iota_{\unit}$ & 0 & 0 & \measlesUnitSpecific &
  0 \\
\hline
\end{tabular}
\caption{Parameters for the measles model. Sub-model {\modelA} has 4 unit-specific parameters and 8 shared parameters, with movement between units following a gravity equation. Sub-model {\modelB} has 12 unit-specific parameters and no shared parameters, with gravity movement. Sub-model {\modelC} has 12 unit-specific parameters and no shared parameters, with independent immigration of infections rather than movement between units. The last column shows the parameter values used for the simulated data in Figure~2 of the main text.
}\label{tab:params}
\end{center}
\end{table}

>>>>

%%% Statistica Sinica formatting
% \newcommand\statSinica[1]{#1}
 \newcommand\statSinica[1]{}
%%% arXiv formatting
 \newcommand\arxiv[1]{#1}
% \newcommand\arxiv[1]{}

@@

The number of travelers from town $\unit$ to $\altUnit$ is denoted by $v_{\unit\altUnit}$.
Here, $v_{\unit\altUnit}$ is constructed using the gravity movement model of \cite{xia04} given by
\[
v_{\unit\altUnit} = \gravity_{\unit} \cdot \frac{\hspace{2mm} \overline{\dist}\hspace{2mm}}{\hspace{1.5mm}\overline{\pop}^2\hspace{1mm}} \cdot \frac{\pop_\unit \cdot \pop_{\altUnit}}{\dist_{\unit\altUnit}},
\]
where $\gravity_u$ is the {\it gravitational constant}, $\dist_{\unit\altUnit}$ is the distance between town $\unit$ and town $\altUnit$, $\pop_\unit$ is the average of $P_{\unit}(t)$ across time, $\overline{\pop}$ is the average of $\pop_{\unit}$ across towns, and $\overline{\dist}$ is the average of $\dist_{\unit\altUnit}$ across all pairs of towns.
The transition processes $N_{EI,\unit}(t)$, $N_{IR,\unit}(t)$ and $N_{QD,\unit}(t)$, for $Q\in\{S,E,I,R\}$,  are modeled as conditional Poisson process with per-capita rates $\mu_{EI,\unit}$, $\mu_{IR,\unit}$ and $\mu_{QD,\unit}$ respectively, and we fix $\mu_{QD,\unit}=(50 \mbox{ year})^{-1}$.
A fraction $\cohort$ of births enter the susceptible cohort on the school admission day, and hence the birth process $N_{BS,\unit}(t)$ is an inhomogeneous Poisson process with rate
%\begin{equation}\nonumber
$\mu_{BS,\unit}(t-\birthdelay) \big[
  (1-\cohort) + \cohort \delta(t-t_a) \big]$,
%\end{equation}
where $\mu_{BS,\unit}(t)$ is specified by interpolated census data, $t_a=\lfloor t \rfloor + 252/365$ is the admission date for the year containing $t$, $\delta$ is a Dirac delta function, and $\birthdelay=4\, \mbox{ year}$ is a fixed delay between birth and entry into a high-transmission school community.

@@

The simulated model is drawn from model {\modelA}, which is nested within model {\modelB} but not within {\modelC}.
For models {\modelB} and {\modelC}, with all parameters are unit-specific, we must estimate $20\times 13=260$ parameters.
This greatly exceeds the 7 shared parameters fitted by \citet{ionides21} for a measles metapopulation model using a large computational effort.
An iterated guided intermediate resampling filter (IGIRF) algorithm was used to fit 9 shared parameters and 3 unit-specific initial value parameters in a measles metapopulation model \citep{park20}.
That analysis used a customized treatment for the unit-specific initial value parameters and did not attempt to estimate other unit-specific parameters.
IGIRF is sensitive to the choice of the guide function, and the model-specific implementation used by \citet{park20} performed better than the generic implementation of IGIRF which is currently available in \code{spatPomp}.
To our knowledge, {\ibpf} is therefore advancing the current limits for scalability of simulation-based maximum-likelihood inference for metapopulation dynamics.

@@

<<sim_search_boxplot, echo=F, fig.height=3, fig.width=5.5, out.width="5in", fig.cap = "Fitting simulated measles data, using an initial search and four refinement steps. (A1--A5) model \\modelA, $4\\times 20$ unit-specific parameters and $9$ shared parameters; (B1--B5) model \\modelB, $13\\times 20$ unit-specific parameters and no shared parameters;  (C1--C5) model \\modelC, also all unit-specific, but with immigration rather than coupling, matching He et al (2010). The horizontal dashed line is the log-likelihood at the true parameters, evaluated by BPF.",fig.show = "hold">>=
a <- cbind(
A1 = v1a$loglik,
A2 = v2a$loglik,
A3 = v3a$loglik,
A4 = v4a$loglik,
A5 = v5a$loglik
)

@@

The MLE is not known exactly in this case.
Wilks' theorem gives an asymptotic expectation that the log-likelihood at the MLE should be greater than the likelihood at the truth by approximately $1/2$ the number of parameters, which here is $(4\times 20+9)/2=44.5$ for {\modelA} and $(13\times 20)/2=130$ for {\modelB}.
The $y$-axis values in Fig.~\ref{fig:sim_search_boxplot} show log-likelihoods exceeding the truth by less than this, indicating some imperfection to the Monte Carlo maximization so far as Wilks' asymptotic result holds in this situation.
Despite this limitation, searches that exceed the likelihood at the truth have found inferentially plausible sets of parameters which can be used to study the likelihood surface around the maximum.
For example, Monte Carlo profile confidence intervals can give proper coverage even in the presence of considerable Monte Carlo error \citep{ionides17,ning21}.

@@

\arxiv{
\vspace{2mm}
\noindent{\bf Version notes}

\noindent Version [v1] was submitted to {\it Statistica Sinica}, and [v2] was accepted for publication. [v3] corrects for an omission of the cohort effect in the description of the model in Section~\ref{sec:model}.
}

@@

\arxiv{
\newpage

\setcounter{section}{0}
\setcounter{equation}{0}
\def\theequation{S\arabic{section}.\arabic{equation}}
%\def\thesection{S\arabic{section}}
\renewcommand\thefigure{S\arabic{figure}}
\renewcommand\thetable{S\arabic{table}}
\def\thesection{S\arabic{section}}

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{|l|c|c|c|c|}
\hline
Parameter & \modelA & \modelB & \modelC & Simulation \\
\hline
$\pinit_{S,u}$ & \measlesUnitSpecific & \measlesUnitSpecific & \measlesUnitSpecific &
  $\Sexpr{myround(theta["S_01"],3)}$ \\
$\pinit_{E,u}$ & \measlesUnitSpecific & \measlesUnitSpecific & \measlesUnitSpecific &
  $\Sexpr{myround(theta["E_01"],5)}$ \\
$\pinit_{I,u}$ & \measlesUnitSpecific & \measlesUnitSpecific & \measlesUnitSpecific &
  $\Sexpr{myround(theta["I_01"],5)}$ \\
$\rho_u$ & \measlesUnitSpecific & \measlesUnitSpecific & \measlesUnitSpecific &
  $\Sexpr{myround(theta["rho1"],1)}$ \\
$\measOD_{\unit}$ & \measlesShared & \measlesUnitSpecific & \measlesUnitSpecific &
  $\Sexpr{myround(theta["psi1"],2)}$ \\  
$\mu_{EI,\unit}$ & \measlesShared & \measlesUnitSpecific & \measlesUnitSpecific &
  $\Sexpr{myround(theta["sigma1"]/52,0)} \, \mathrm{week}^{-1}$ \\
$\mu_{IR}$ & \measlesShared & \measlesUnitSpecific & \measlesUnitSpecific &
  $\Sexpr{myround(theta["gamma1"]/52,0)} \, \mathrm{week}^{-1}$ \\
$\meanBeta_\unit$ & \measlesShared & \measlesUnitSpecific & \measlesUnitSpecific &
  $\Sexpr{myround(theta["gamma1"]*(theta["R01"]+theta["mu1"])/52,0)} \, \mathrm{week}^{-1}$ \\
$\sigma_{SE,\unit}$ & \measlesShared & \measlesUnitSpecific & \measlesUnitSpecific &
  $\Sexpr{myround(theta["sigmaSE1"],2)} \, \mathrm{year}^{1/2}$ \\
$\amplitude_{\unit}$ & \measlesShared & \measlesUnitSpecific & \measlesUnitSpecific &
  $\Sexpr{myround(theta["amplitude1"],1)}$ \\
$\alpha_{\unit}$ & \measlesShared & \measlesUnitSpecific & \measlesUnitSpecific &
  $\Sexpr{myround(theta["alpha1"],0)}$ \\
$\cohort$& \measlesShared & \measlesUnitSpecific & \measlesUnitSpecific &
  $\Sexpr{myround(theta["cohort1"],0)}$ \\  
$\gravity_{\unit}$ & \measlesShared & \measlesUnitSpecific & 0 &
  $\Sexpr{myround(theta["g1"],0)}$ \\
$\iota_{\unit}$ & 0 & 0 & \measlesUnitSpecific &
  0 \\
\hline
\end{tabular}
\caption{Parameters for the measles model. Sub-model {\modelA} has 4 unit-specific parameters and 9 shared parameters, with movement between units following a gravity equation. Sub-model {\modelB} has 13 unit-specific parameters and no shared parameters, with gravity movement. Sub-model {\modelC} has 13 unit-specific parameters and no shared parameters, with independent immigration of infections rather than movement between units. The last column shows the parameter values used for the simulated data in Figure~2 of the main text.
}\label{tab:params}
\end{center}
\end{table}
